A tool set for random number generation on GPUs in R

### 2022-01-10 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.tex
update paper 1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@3108 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Below we generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{rnormGpu}. The best performance we have seen using \fct{rnormGpu} is above over 100 times faster with $512 \times 128$ work-items than using the standard (and single-threaded) \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size increases.
\begin{CodeChunk}
\begin{CodeInput}
R> streams <- createStreamsGpu(512 * 128)
R> system.time(rnormGpu(c(10000, 10000), streams = streams, Nglobal = c(512,
+    128), type = "double"))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   0.044   0.004   0.047
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> streams <- createStreamsGpu(n = 256*64)
R> month_gpu <- vclMatrix(month, type = "integer")
R> system.time(result_month <- 
+  clrng::fisher.sim(month_gpu, 1e6, 
+     streams=streams, type="double", 
+     returnStatistics=TRUE,  Nglobal = c(256,64)))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   0.314   0.000   0.316
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> week_GPU <- gpuR::vclMatrix(week, type = "integer")
R> system.time(resultWeek <- clrng::fisher.sim(week_GPU, 10000000,
+    streams = streams, type = "double", returnStatistics = TRUE,
+    Nglobal = c(256, 64)))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   1.926   0.016   1.947
\end{CodeOutput}
\begin{CodeInput}
R> unlist(resultWeek[c("threshold", "simNum", "counts")])
\end{CodeInput}
\begin{CodeOutput}
## threshold    simNum    counts 
##    -54990  10010624      1350
\end{CodeOutput}
\begin{CodeInput}
R> resultWeek$p.value
\end{CodeInput}
\begin{CodeOutput}
## [1] 0.000135
\end{CodeOutput} 
\end{CodeChunk}

@@

%Due to the importance of Gaussian random field, there are a large amount of studies devoted to Gaussian random field generation techniques. 
\cite{LiuandLi2019} gives a comprehensive review on seven popular methods for Gaussian random field generation, %which are the turning bands method \citep{matheron_1973}, spectral method \citep{Meja1974OnTS,1972JSV}, matrix decomposition method \citep{davis1987}, Karhunen-Lo\'eve expansion \citep{LoeveM:1978}, moving average method \citep{journel1974, oliver1995moving}, sequential simulation \citep{johnson1987multivariate, gomez1993joint, pebesma2004multivariable}, and local average subdivision \citep{fenton1990simulation}. 
all of these methods except the direct matrix decomposition method, are approximations to the random field and have specific requirements on the type of grid or covariance functions. The direct matrix decomposition method is exact, it works for all covariance functions and can generate random fields on arbitrary grids, and is straightforward to implement \citep{LiuandLi2019}. Other \proglang{R} packages that offer simulation of Gaussian random fields such as \pkg{geoR} \citep{geoR2001}, does not work for large number of locations. The \pkg{RandomFields} \citep{RandomFields2015} package has several different methods for simulation of Gaussian fields, among which the circulant embedding, which is an improved method on direct matrix decomposition, and some variants of the method like the cut-off embedding \citep{gneiting2006fast} are also exact and fast for isotropic Gaussian random fields, however, they work only on rectangular grids. We are able to simulate on GPU exact Gaussian fields on arbitrary grids by using the direct matrix decomposition method. The implementation of this method is as follows.

@@

% $U=L*D^{1/2}*Z$, where 
% \begin{itemize}
% \itemsep0em 
% \item $L$ is the lower unit triangular matrix in Cholesky decomposition of $\Sigma = L*D*L^\top$,
% \item $D$ is a diagonal matrix.
% \item $Z=(Z_1, Z_2, \dots, Z_n) \sim \text{MVN}(0,I_n)$, where $I_n$ is a $n \times n$ identity matrix.
% \end{itemize}
The downside of this method is that it's computationally expensive and is thus very slow. When the number of locations $n$ is large, the time cost of Cholesky decomposition of the covariance matrix is $\mathcal{O}(n^3)$,  and is $\mathcal{O}(n^2)$ for the matrix-vector multiplication $L*Z$ \citep{LiuandLi2019}.  Our other \proglang{R} package \pkg{gpuBatchMatrix} is able to address this issue since it computes batches of Mat\'ern covariance matrices in parallel, and does Cholesky decomposition and matrix-matrix multiplication in batches in parallel on GPU. The following shows an example, in which we simulate on GPU eight Gaussian random fields of Mat\'ern covariance at one time with four sets of parameters, by taking advantage of the GPU-capabilities offered by \pkg{clrng} and \pkg{gpuBatchMatrix} together.
%\pkg{clrng} package not only does the matrices decomposition (Cholesky decomposition and matrix-vector multiplication) in parallel, but also computes the covariance matrices in parallel on GPU.

\subsection{Simulating Gaussian random fields with Mat\'ern covariances}
\begin{CodeChunk}
\begin{CodeInput}
R> library("gpuBatchMatrix")
R> library("geostatsp")
\end{CodeInput} 
\end{CodeChunk}

>>>>

Below we generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{rnormGpu}. The best performance we have seen using \fct{rnormGpu} is above over 100 times faster with $512 \times 128$ work-items than using the standard (and single-threaded) \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size increases.
\begin{CodeChunk}
\begin{CodeInput}
R> streams <- createStreamsGpu(512 * 128)
R> system.time(rnormGpu(c(10000, 10000), streams = streams, Nglobal = c(512,
+    128), type = "double"))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   0.045   0.004   0.049
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> streams <- createStreamsGpu(n = 256*64)
R> month_gpu <- vclMatrix(month, type = "integer")
R> system.time(result_month <- 
+  clrng::fisher.sim(month_gpu, 1e6, 
+     streams=streams, type="double", 
+     returnStatistics=TRUE,  Nglobal = c(256,64)))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   0.319   0.000   0.319
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> week_GPU <- gpuR::vclMatrix(week, type = "integer")
R> system.time(resultWeek <- clrng::fisher.sim(week_GPU, 10000000,
+    streams = streams, type = "double", returnStatistics = TRUE,
+    Nglobal = c(256, 64)))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   1.923   0.023   1.949
\end{CodeOutput}
\begin{CodeInput}
R> unlist(resultWeek[c("threshold", "simNum", "counts")])
\end{CodeInput}
\begin{CodeOutput}
## threshold    simNum    counts 
##    -54990  10010624      1350
\end{CodeOutput}
\begin{CodeInput}
R> resultWeek$p.value
\end{CodeInput}
\begin{CodeOutput}
## [1] 0.000135
\end{CodeOutput} 
\end{CodeChunk}

@@

%Due to the importance of Gaussian random field, there are a large amount of studies devoted to Gaussian random field generation techniques. 
\cite{LiuandLi2019} gives a comprehensive review on seven popular methods for Gaussian random field generation, %which are the turning bands method \citep{matheron_1973}, spectral method \citep{Meja1974OnTS,1972JSV}, matrix decomposition method \citep{davis1987}, Karhunen-Lo\'eve expansion \citep{LoeveM:1978}, moving average method \citep{journel1974, oliver1995moving}, sequential simulation \citep{johnson1987multivariate, gomez1993joint, pebesma2004multivariable}, and local average subdivision \citep{fenton1990simulation}. 
all of these methods except the direct matrix decomposition method, are approximations to the random field and have specific requirements on the type of grid or covariance functions. The direct matrix decomposition method is exact, it works for all covariance functions and can generate random fields on arbitrary grids, and is straightforward to implement \citep{LiuandLi2019}. By using this method, we are able to simulate on GPU exact Gaussian fields on arbitrary grids. The implementation of this method is as follows.

@@

The downside of this method is that it's computationally expensive and is thus very slow. When the number of locations $n$ is large, the time cost of Cholesky decomposition of the covariance matrix is $\mathcal{O}(n^3)$,  and is $\mathcal{O}(n^2)$ for the matrix-vector multiplication $L*Z$ \citep{LiuandLi2019}. Methods employing approximations gain benefits in computations while sacrificing exactness, we preserve the exactness and also largely reduce the computation burden by utilizing our other \proglang{R} package \pkg{gpuBatchMatrix}. \pkg{gpuBatchMatrix} address this issue since it computes batches of Mat\'ern covariance matrices in parallel, and does Cholesky decomposition and matrix-matrix multiplication in batches in parallel on GPU.

Other \proglang{R} packages that offer simulation of Gaussian random fields such as \pkg{geoR} \citep{geoR2001}, does not work for large number of locations. The \pkg{RandomFields} \citep{RandomFields2015} package has several different methods for simulation of Gaussian fields, among which the circulant embedding, which is an improved method on direct matrix decomposition, and some variants of the method like the cut-off embedding \citep{gneiting2006fast} are also exact and fast for isotropic Gaussian random fields, however, they work only on rectangular grids.

% $U=L*D^{1/2}*Z$, where 
% \begin{itemize}
% \itemsep0em 
% \item $L$ is the lower unit triangular matrix in Cholesky decomposition of $\Sigma = L*D*L^\top$,
% \item $D$ is a diagonal matrix.
% \item $Z=(Z_1, Z_2, \dots, Z_n) \sim \text{MVN}(0,I_n)$, where $I_n$ is a $n \times n$ identity matrix.
% \end{itemize}
%\pkg{clrng} package not only does the matrices decomposition (Cholesky decomposition and matrix-vector multiplication) in parallel, but also computes the covariance matrices in parallel on GPU.

\subsection{Simulating Gaussian random fields with Mat\'ern covariances}
The following shows an example, in which we simulate on GPU eight Gaussian random fields of Mat\'ern covariance at one time with four sets of parameters, by taking advantage of the GPU-capabilities offered by \pkg{clrng} and \pkg{gpuBatchMatrix} together.
\begin{CodeChunk}
\begin{CodeInput}
R> library("gpuBatchMatrix")
R> library("geostatsp")
\end{CodeInput} 
\end{CodeChunk}
